---
title: "Market Regime Detection in Chilean Equities"
subtitle: "Hidden Markov Models on the IPSA, mapped to structural-break events"
author: "MIEL"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 12,
  fig.height = 8
)

# ---- Libraries (loaded once) ----
library(tidyverse)   # dplyr, tidyr, ggplot2, purrr, stringr, forcats, ...
library(tidyquant)   # Yahoo Finance price download
library(depmixS4)    # Gaussian Hidden Markov Models
library(lubridate)   # date handling
library(scales)      # axis formatting
library(zoo)         # rolling statistics
library(slider)      # sliding-window helpers

# depmixS4 (via MASS/nlme) masks several dplyr verbs; pin them back to dplyr
select    <- dplyr::select
filter    <- dplyr::filter
summarise <- dplyr::summarise

set.seed(42)         # reproducible k-means

# Analysis window and output location for the exported figure
START_DATE  <- as.Date("2018-01-01")
FIG_DIR     <- "figures"
if (!dir.exists(FIG_DIR)) dir.create(FIG_DIR, recursive = TRUE)
```

## 1. IPSA constituents

The universe is the ~28 constituents of the IPSA (Santiago Stock Exchange's
main index). Yahoo Finance suffixes Chilean tickers with `.SN`.

```{r tickers}
ipsa_tickers <- tibble::tibble(
  asset = c(
    "Aguas Andinas", "Embotelladora Andina", "Banco de Crédito e Inversiones",
    "Banco Santander Chile", "CAP", "Compañía Cervecerías Unidas",
    "Cencosud Shopping", "Cencosud", "Banco de Chile", "Empresas CMPC",
    "Colbún", "Viña Concha y Toro", "Empresas Copec", "Engie Energía Chile",
    "Enel Américas", "Enel Chile", "Entel Chile", "S.A.C.I. Falabella",
    "Inversiones Aguas Metropolitanas", "Itaú Corpbanca", "Mallplaza",
    "Parque Arauco", "Ripley", "Grupo Security", "SMU", "Sonda",
    "Sociedad Química y Minera de Chile (SQM-B)",
    "Compañía Sudamericana de Vapores"
  ),
  ticker = c(
    "AGUAS-A.SN", "ANDINA-B.SN", "BCI.SN", "BSANTANDER.SN", "CAP.SN",
    "CCU.SN", "CENCOSHOPP.SN", "CENCOSUD.SN", "CHILE.SN", "CMPC.SN",
    "COLBUN.SN", "CONCHATORO.SN", "COPEC.SN", "ECL.SN",
    "ENELAM.SN", "ENELCHILE.SN", "ENTEL.SN", "FALABELLA.SN",
    "IAM.SN", "ITAUCORP.SN", "MALLPLAZA.SN", "PARAUCO.SN", "RIPLEY.SN",
    "SECURITY.SN", "SMU.SN", "SONDA.SN", "SQM-B.SN", "VAPORES.SN"
  )
)
```

## 2. Download prices and compute log returns

For each ticker, pull daily adjusted prices from `START_DATE` to today and
compute log returns. Series that are too short or missing an `adjusted`
column are skipped.

```{r download}
get_asset_returns <- function(ticker, asset) {
  tryCatch({
    df <- tq_get(ticker, from = START_DATE)

    if (nrow(df) < 500 || !"adjusted" %in% names(df)) {
      message(sprintf("Skipping %s (%s): too short or missing 'adjusted'",
                      asset, ticker))
      return(NULL)
    }

    df %>%
      arrange(date) %>%
      mutate(
        log_return = log(adjusted / lag(adjusted)),
        asset      = asset
      ) %>%
      drop_na()

  }, error = function(e) {
    message(sprintf("Error loading %s (%s): %s", asset, ticker, e$message))
    NULL
  })
}

returns_list <- purrr::map2(ipsa_tickers$ticker, ipsa_tickers$asset,
                            get_asset_returns)
prices <- bind_rows(purrr::compact(returns_list))
```

## 3. Per-asset regimes (Hidden Markov Model)

Fit a 5-state Gaussian HMM to each stock's return series and Viterbi-decode
its latent state path.

```{r hmm}
fit_hmm_depmix <- function(ts_data, nstates = 5) {
  df  <- data.frame(y = ts_data)
  mod <- depmix(y ~ 1, data = df, nstates = nstates, family = gaussian())
  fit <- fit(mod, verbose = FALSE)
  posterior(fit, type = "viterbi")$state
}

regime_results <- prices %>%
  group_by(asset) %>%
  group_modify(~ {
    ts_data    <- .x$log_return
    asset_name <- unique(.x$asset)

    # Guard against non-finite or degenerate series
    if (any(!is.finite(ts_data)) || length(unique(ts_data)) < 10) {
      message(sprintf("Skipping %s: non-finite or too-short series", asset_name))
      return(tibble(date = .x$date, asset = asset_name, regime = NA))
    }

    tryCatch({
      decoded <- fit_hmm_depmix(ts_data, nstates = 5)
      tibble(date = .x$date, asset = asset_name, regime = decoded)
    }, error = function(e) {
      message(sprintf("Error fitting %s: %s", asset_name, e$message))
      tibble(date = .x$date, asset = asset_name, regime = NA)
    })
  }) %>%
  ungroup()
```

## 4. Harmonize regimes across assets (k-means)

Raw HMM states aren't comparable across stocks, so each (asset, state) is
summarized by its mean return and volatility, then k-means-clustered into
five interpretable, cross-asset regime types.

```{r cluster}
regime_stats <- regime_results %>%
  left_join(prices %>% select(date, asset, log_return),
            by = c("date", "asset")) %>%
  filter(!is.na(log_return), !is.na(regime)) %>%
  group_by(asset, regime) %>%
  summarise(
    mean_return = mean(log_return),
    sd_return   = sd(log_return),
    n_days      = n(),
    .groups     = "drop"
  ) %>%
  filter(n_days >= 10)   # drop unstable / short-lived states

# Cluster on standardized (mean, volatility)
clustering_input <- regime_stats %>% mutate(row_id = row_number())
clustering_data  <- clustering_input %>% select(mean_return, sd_return) %>% scale()

km <- kmeans(clustering_data, centers = 5)

regime_stats <- regime_stats %>%
  mutate(row_id = row_number()) %>%
  left_join(
    clustering_input %>%
      mutate(regime_type_cluster = km$cluster) %>%
      select(row_id, regime_type_cluster),
    by = "row_id"
  ) %>%
  select(-row_id)

# Label each cluster by its ACTUAL return/volatility profile. k-means numbers
# clusters arbitrarily, so a fixed 1:5 -> name mapping is meaningless (it made
# the whole heatmap collapse to one color). Instead, derive each label from the
# cluster's day-weighted mean return and volatility, deterministically.
cluster_profile <- regime_stats %>%
  group_by(regime_type_cluster) %>%
  summarise(
    m = weighted.mean(mean_return, n_days),
    s = weighted.mean(sd_return,   n_days),
    .groups = "drop"
  )

assign_regime_labels <- function(prof) {
  prof$regime_type <- NA_character_
  prof$regime_type[which.min(prof$m)] <- "Bearish Shock"    # sharpest drawdown
  prof$regime_type[which.max(prof$m)] <- "Bullish Strong"   # strongest rally
  rest <- which(is.na(prof$regime_type))
  vn <- rest[which.max(prof$s[rest])]                        # choppiest of the rest
  prof$regime_type[vn] <- "Volatile Neutral"
  rest <- which(is.na(prof$regime_type))
  ord <- rest[order(prof$m[rest])]                           # remaining two by mean
  prof$regime_type[ord[1]] <- "Bearish Stable"
  prof$regime_type[ord[2]] <- "Bullish Stable"
  prof
}

cluster_labels <- assign_regime_labels(cluster_profile) %>%
  select(regime_type_cluster, regime_type)

regime_stats <- regime_stats %>%
  left_join(cluster_labels, by = "regime_type_cluster")

regime_labeled <- regime_results %>%
  left_join(regime_stats %>% select(asset, regime, regime_type),
            by = c("asset", "regime"))
```

## 5. Market-wide switch detection

Count how many assets change regime each day and flag co-movement spikes as
candidate structural-break dates.

```{r switches}
switch_counts <- regime_results %>%
  group_by(asset) %>%
  arrange(date) %>%
  mutate(switched = regime != lag(regime)) %>%
  ungroup() %>%
  filter(switched, !is.na(switched)) %>%
  count(date, name = "n_switches")

# Local peaks where the daily switch count is both a local max and large
switch_peaks <- switch_counts %>%
  arrange(date) %>%
  mutate(local_max = zoo::rollapply(
    n_switches, 11,
    function(x) {
      center <- ceiling(length(x) / 2)
      x[center] == max(x) & x[center] >= 8
    },
    fill = NA, align = "center"
  )) %>%
  filter(local_max)

switch_dates <- switch_peaks$date
cat("Detected candidate regime-change dates:\n")
print(as.character(switch_dates))
```

## 6. Final annotated heatmap

Asset × time regime heatmap. Dashed lines mark the detected structural-break
dates, annotated with the real political / economic events they coincide with.
The figure is exported to `figures/regime_heatmap.png`.

```{r final-plot}
# Structural breaks: the dates the detector flags at >= 8 simultaneous switches,
# each annotated with the real event it coincides with (alternating label height).
switch_labels <- tibble::tibble(
  date = as.Date(c(
    "2019-10-21", "2019-11-18", "2020-03-12",
    "2021-05-17", "2021-11-23", "2024-08-05"
  )),
  label = c(
    "Estallido Social",
    "Acuerdo Constitucional",
    "COVID-19",
    "Convención Constituyente",
    "Presidencial & Parlamentaria",
    "Yen carry unwind"
  ),
  above = c(TRUE, FALSE, TRUE, FALSE, TRUE, FALSE)
)

# Specification-dependent finding: 2024-10-17 (Banco Central de Chile rate cut)
# only clears the threshold under a 4-state HMM, not this 5-state one. Shown as a
# distinct dotted marker to make the finding legible on the figure.
finding_label <- tibble::tibble(
  date  = as.Date("2024-10-17"),
  label = "BCCh rate cut · n=4 only",
  above = FALSE
)

# Place annotations just above / below the plotted assets (count varies by run,
# since some tickers occasionally fail to download from Yahoo)
n_assets <- dplyr::n_distinct(regime_labeled$asset)
switch_labels  <- switch_labels  %>% mutate(y = ifelse(above, n_assets + 1.5, -1.5))
finding_label  <- finding_label  %>% mutate(y = ifelse(above, n_assets + 1.5, -1.5))

regime_plot <- ggplot(regime_labeled,
                      aes(x = date, y = fct_rev(asset), fill = regime_type)) +
  geom_tile(height = 0.8) +
  scale_fill_manual(
    values = c(
      "Bullish Stable"   = "#4CAF50",
      "Volatile Neutral" = "#FF9800",
      "Bearish Stable"   = "#2196F3",
      "Bearish Shock"    = "#F44336",
      "Bullish Strong"   = "#9C27B0"
    ),
    na.value = "gray80"
  ) +
  geom_vline(data = switch_labels, aes(xintercept = date),
             color = "black", linetype = "dashed", linewidth = 0.9) +
  geom_text(
    data = switch_labels,
    aes(x = date, y = y, label = label),
    size = 3, fontface = "bold",
    vjust = ifelse(switch_labels$above, 0, 1),
    inherit.aes = FALSE
  ) +
  # specification-dependent finding (dotted, grey)
  geom_vline(data = finding_label, aes(xintercept = date),
             color = "grey30", linetype = "dotted", linewidth = 0.9) +
  geom_text(
    data = finding_label,
    aes(x = date, y = y, label = label),
    size = 2.7, fontface = "italic", color = "grey30",
    vjust = ifelse(finding_label$above, 0, 1),
    inherit.aes = FALSE
  ) +
  labs(
    title    = "Cluster de cambio de régimen (Hidden Markov Model) de acciones IPSA",
    subtitle = "Líneas punteadas indican fechas de cambio estructural percibido",
    x = NULL, y = "Asset", fill = "Tipo de Régimen"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x     = element_text(angle = 45, hjust = 1),
    axis.text.y     = element_text(size = 8),
    legend.position = "bottom",
    plot.margin     = margin(30, 10, 60, 10)
  ) +
  coord_cartesian(clip = "off")

# Save to disk and also render inline
ggsave(
  filename = file.path(FIG_DIR, "regime_heatmap.png"),
  plot     = regime_plot,
  width    = 12, height = 8, dpi = 200, bg = "white"
)

regime_plot
```
